\textbf{ Spectrophotometric parallaxes with linear models:\\ Accurate distances for luminous red-giant stars }

### 2018-08-13 paper/sp.tex
words

% To-do items
% -----------
% - make sure we address the sensibleness and generality of the linear model for the photometric space
% - Coryn's point that we should carefully explain how our model is geometric.
% - related: HWR's point that this is standardizeable candles, like the Ia SNe results.
% - related: Put in an explanation of WHY this works so well.
% - related: Put in an explanation of how it is that we aren't really modeling luminosity or distance but really just generating parallaxes.
% - does the model need a cute name, like The Swan?
% - check with WISE experts that we have the correct WISE photometry
% - general HWR point that we aren't putting in to the model any idea about how apparent magnitude varies with distance; this means that we rely on data for EVERYTHING.
% - write words about the setting of Lambda and the parallax offset (currently 0.0483 +/- 0.012 mas)
% - we have shown that outliers in sp - gaia tend to have large H-W2 or excess flux like in crowded regions, but there are some exceptions. Our error estimates are consistent with no intrinsic scatter if we use a robust chi-squared estimate. That is, chi-squared is dominated by outliers. We clipped the bad apogee spectral pixels at an error of 0.05 in normalized flux. Our uncertainty estimates are sensitive to this. The reader is asked to suck it.
% - write words about other linear models we have made or could make, probably in the discusssion.
% - are we citing all the proper DR2 papers we need to cite?

@@

\title{\textbf{%
Spectrophotometric parallaxes with \project{The Cygnet}:\\
Accurate distances for luminous red-giant stars
}}

@@

\begin{abstract}\noindent
With contemporary infrared spectroscopic surveys like \apogee,
red-giant stars can be observed to distances and extinctions
at which geometric parallaxes are not precisely measured by \gaia.
Here we use the \apogee--\gaia\ overlap to train a purely linear model of
continuum-normalized \apogee\ spectroscopy
(plus \gaia, \zmass, and \wise\ photometry)
that predicts distance modulus (logarithmic parallax or distance)
for red stars with $0<\logg<2.2$.
The training makes use of a justified likelihood function for the \gaia\ parallaxes,
so that the training can be performed with full use of uncertainties and
without any cuts on parallax or parallax signal-to-noise ratio (not even any non-negative
requirements).
There are no physical inputs to the model; it is a pure regression designed
to accurately predict astrometric parallaxes.
The model includes an L1 regularization that zeros out the contributions of
most spectral pixels to the parallax estimates.
The training is performed
with leave-out subsamples such that no star's astrometry is used even indirectly in its
spectrophotometric parallax estimate.
The model is flexible enough---and has enough photometric information---to
correct for extinction by dust without any
input of explicit extinction or reddening estimates.
We re-label each star in the sample
with a new spectrophotometric parallax; this new parallax has an empirical (1-sigma)
uncertainty of 9~percent, which is more precise than the \gaia\ parallax
for the vast majority of targets.
Validation with globular and open clusters shows that the spectrophotometric parallaxes
are both accurate and precise.
These distance estimates open up great opportunities for
mapping the Milky Way disk with the next
generation of spectroscopic surveys, and especially \sdssv.
\end{abstract}

@@

If we want to make precise kinematic and element-abundance maps
of the Milky Way disk out to large heliocentric radii,
and especially through the extinction to the other side of the Galactic Center, we will
need to make use of luminous red giants, and we will need to observe in
the infrared.
These arguments motivate the \apogee\ (CITE), \apogee-2 (CITE), and \sdssv\ (\citealt{sdssv})
projects, which take spectra in the infrared, and deliver detailed abundances
along the entire red-giant branch up to the tip.
The maps made with these data will reveal the dynamics of the Milky Way and its disk,
and show us how and where stars form, and how they migrate around (and out of) the
disk with time.

As we operate these spectroscopic surveys,
the \gaia\ Mission (CITE) has been revolutionizing our view of the disk.
It delivers good proper motions everywhere,
and extremely valuable parallax information
locally.
However, it does not deliver precise geometric parallaxes over a large fraction of
the disk, and certainly not in the dusty and crowded regions.
For these reasons, the value of \gaia\ in these infrared spectroscopy projects is not
to deliver distance information directly, but rather to calibrate stellar models,
which then deliver distance information through relationships between stellar
luminosities and spectral characteristics.

Luminous red-giant stars are valuable for mapping the Milky Way disk for a number
lf reasons, only one of which is their great luminosities (and hence brightnesses
even at large distance).
They are extremely common (frequent) stars, much more common than distance-indicating
stars like Cepheids.
They are produced by stellar populations at all metallicities and almost all ages,
so they can be found in all parts of the Galaxy.
They have temperatures and surface gravities such that it is
straightforward to spectroscopically measure metallicities,
detailed chamical abundances, and stellar parameters, including ages (\citealt{martig, nessage}).
The red-giant branch is photometrically near-orthogonal to reddening vectors by dust,
so they are easy to de-redden or dust-correct.
And finally, in physical models of red giants, their predicted luminosities are simple
functions of composition, surface gravity, temperature, and age.
Thus, if we can measure these properties of red giants, we can (in principle) predict
their luminosities and use them as standard candles.
That is---like type Ia supernovae---red giants are standardizable candles.

In this project, we develop purely data-driven techniques to predict
red-giant parallaxes or distances with spectroscopy and photometry.
Our method capitalizes on the point that red-giant stars are dust-correctable,
standardizable candles, but it makes no use whatsoever of any stellar interior,
stellar exterior, or dust-reddening models.
We only use these physical expectations about red giants to recommend
data features---inputs---for the data-driven model.
The data-driven method uses patterns in the observed data themselves
to discover the relationships between spectral features in the spectra of the stars,
photometry (including colors), and parallax (or distance).

@@

Simple data-driven models (like the ones we will employ here)
suffer from many limitations for this kind of work:
Because they are constrained only by the data, and not physical law,
they can learn relationships that we know
to be physically unlikely, and (by the same token) they can't benefit from our
physical knowledge.
They are required, in some sense, to spend some of the information in the data on
learning physical laws or relationships.
Not all of the information in the data is being used
optimally when we ignore our physical models.
Relatedly, they are not (usually) interpretable or generalizable:
A data-driven model trained with one kind of data can not usually be used in another
context with very different input data.
And the internal properties of the model are not useful, in general, for improving
our understanding or intuition about the astrophysics in play.

@@

What follows is a regression:
We use \gaia\ measurements of parallax to learn relationships between 
the spectroscopic and photometric properties of stars and their parallaxes.
We then use that learned model to estimate improved parallaxes for stars with
poor (or no) \gaia\ measurements.
This regression is very sensitive to certain aspects of the data or experimental
design..
One is that we will do far better with the regression if we have an accurate
model for noise process in the \gaia\ data.
For this reason, we make use of a justifiable likelihood function for the
\gaia\ astrometry (\citealt{gaialf}).
Another is that regressions can be strongly biased if we have a bad selection function
or data censoring.
This means that we cannot apply cuts to the \gaia\ astrometry or signal-to-noise
unless those cuts are correctly accounted for in our generative model for the
data sample.
Below, we will make no such cuts. That means that our training step has the odd
property that much of the training data is low in signal-to-noise and many of the
training parallax measurements are even negative!
The fact that we are using a justified likelihood function ensures that these
data will only affect the model in sensible and justifiable ways.
The alternative---cutting to high signal-to-noise data---is the more traditional
approach, but it leads to biased inferences (unless the cuts are included correctly
in the likelihood function).
The method in which we make no cuts and keep an umodified, justified likelihood function
is unbiased, and simple.

@@

Our position is that a methodological technique is correct inasmuch as
it delivers correct or best results \emph{under its particular assumptions}.
In that spirit, we present here the assumptions of the method
explicitly.
We will return to these assumptions in the Discussion section below,
to address the costs and benefits of each in more depth.
\begin{description}
\item[features] Perhaps our most fundamental assumption is that the parallax
of a star can be predicted from the features we provide, which are
the full set of (pseudo-continuum-normalized) pixels from the \apogee\ spectrum,
plus the $G$ and $(\BP-\RP)$ photometry from \gaia,
plus the $J$, $H$, and $K_s$ photometry from \zmass\ (\citealt{zmass}),
plus the $W_1$ and $W_2$ photometry from \wise.
That is, we assume that these spectrophotometric \emph{features} are sufficient
to predict the parallax in the face of variations in stellar
age, evolutionary phase, composition, and other parameters, and also interstellar
extinction.

@@

\item[linearity] Perhaps the most restrictive assumption of this work is that
the logarithmic parallax (or, equivalently, the distance modulus) 
can be predicted as a \emph{completely linear function} of
the chosen features. We are only assuming this linearity in a small range of stellar
surface gravity $\logg$, but this assumption is strong, and limits strongly the
flexibility or capacity of the model.
We make it to ensure that our method is easy to optimize, and the results are easy
to interpret.

\item[likelihood] We assume that the \gaia\ parallaxes are generated by a particular
stochastic process, in which the difference between the \gaia-Catalog parallax (plus
a small \gaia-recommended offset; \citealt{lindegren}) and the true parallax is effectively drawn from a
Gaussian with a width set by the \gaia-Catalog uncertainty on the parallax.
This is the standard assumption in all properly probabilistic \gaia-Mission inferences
to date, but it subsumes a number of related assumptions, like that the \gaia\ noise
model is correct, that the stars are only covariant at negligible levels, and that
there are no significant outliers or large-scale systematics.
\end{description}

@@

In addition, we assume that many entries in the parameter $D$-vector $\theta$ will be zero
(the ``sparsity'' assumption).
In order to encourage this, we optimize not $\chi^2$ but a regularized objective function
\begin{eqnarray}
\hat{\theta} &\leftarrow& \argmin_{\theta}\left[\frac{1}{2}\,\chi^2(\theta) + \lambda\,||P\cdot\theta||_1^1\right]
\label{eq:objective}\quad ,
\end{eqnarray}
where
$\lambda$ is a regularization parameter,
$P$ is a projection operator that selects out from the $\theta$ vector only those components
that pertain to the \apogee\ spectral pixels,
and $||y||_1^1$ is the L1-norm or sum of absolute values of the components of $y$.
This kind of regularization adds a convex term to the optimization and leads to
sparse optima.
The $P$ operator makes it such that we only regularize the components of $\theta$ that multiply
the spectral pixels; we ask for the spectral model to be sparse but we don't ask for the photometric
model to be sparse.
We set the value of $\lambda$ by a 
very coarse cross-validation.

This optimization is not convex---there are multiple optima in general.
In particular, there is a large, degenerate, pathological optimum where
the exponential function underflows, all parallaxes are predicted to vanish,
and there is no gradient of anything with respect to the parameter $D$-vector $\theta$.
This large, bad optimum (it is a local, not global, optimum) must be avoided in optimization.
In practice we avoid it by optimizing first for the very highest signal-to-noise
(in astrometric parallax) Training Set stars, and then use that first optimum as a
good first guess or initialization for the full optimization.
This method works because in the limit of high signal-to-noise, the problem asymptotically approaches
the convex problem of L1-regularized linear least-square fitting.
Optimization is performed with \code{scipy.optimize} \acronym{L-BFGS-B}.

Once the model is optimized, the output of the model is a prediction of the parallax,
or really what we will call the \emph{spectrophotometric parallax}.
This spectrophotometric parallax
$\sparallax_m$ for any star $m$ in the validation or test set is
assigned according to
\begin{eqnarray}
\sparallax_m &\leftarrow& \exp(\hat{\theta}\cdot x_m)
\quad ,
\end{eqnarray}
where
$\hat{\theta}$ is the optimal parameter vector according
to \equationname~(\ref{eq:objective}),
and
$x_m$ is the feature $D$-vector for star $m$.
Thus the model can be trained on a training set of stars and applied to make
predictions for a validation set or a test set.
The only requirements are that every test-set object $m$  must have a full feature
vector $x_m$ just as every training-set object $n$ must have a full feature
vector $x_n$.

One final detail: We perform all of the fitting in a two-fold train-and-test framework,
in which the data are split into two disjoint subsets, A and B.
The model trained on the A data is used to predict or produce
the spectrophotometric parallaxes $\sparallax_m$ values (and hence
the distances and parallaxes) of the B data,
and the model trained on the B data is used to 
produce the $\sparallax_m$ values of the A data.
This ensures that, in the estimate of any individal star's
parallax, none of the \gaia\ data pertaining to that star were used.
This makes the parallax estimates from the spectrophotometric feature vectors
$x_m$ statistically independent of the \gaia\ parallax measurements.
They are not independent globally---\gaia\ data were used to train the model---but
each star's spectrophotometric parallax estimate is independent
of the astrometric estimate from \gaia\ on a star-by-star basis.

@@

We need complete feature-vector information for all stars.  For this
reason, we define the Parent Sample to be the set of all stars that
meet the \apogee\ and \gaia\ cuts and also
have the complete set of photometry: $G_n$, $\BP_n$, $\RP_n$, $J_n$,
$H_n$, $K_n$, $W_{1n}$, and $W_{2n}$ and also an \apogee\ spectrum.
In addition, we applied two light color cuts to remove stars with obviously
contaminated or outlying photometry:
\begin{eqnarray}
(J_n - K_n) &<& (+0.4\,\mathrm{mag}) + 0.45 * (\BP_n - \RP_n)
\\
(H_n - W_{2n}) &>& (-0.05\,\mathrm{mag})
\quad .
\end{eqnarray}
These cuts removed roughly 2~percent of the \apogee\ stars.
This Parent Sample contains 44808 stars and is shown in \figurename~\ref{fig:parent}.
\begin{figure}
\includegraphics[width=\textwidth]{parent_sample.pdf}
\caption{Color--color diagrams for the Parent Sample stars.
  The points are colored by
  the \apogee\ pipeline estimates of surface gravity.
  The color cuts used to trim outliers are shown as grey lines.\label{fig:parent}}
\end{figure}

Every Parent-Sample star gets, in addition, a randomly assigned binary
label (A or B).
This is used for two-fold validation and jackknife.
In short, we will use the model trained on the A data to assign luminosities
and distances to the B data and \foreign{vice versa}.

For training, we need A and B Training Sets.
We define the Training Set stars to be Parent-Sample stars that also
have a measured \gaia\ parallax.
The Training-Set stars, in addition to meeting all Parent-Sample cuts,
also must meet ACE WHAT ADDITIONAL CUTS to ensure that the parallax
measurements are good.
But---as we have emphasized above---we do not cut ever on parallax or
parallax error or their ratio. For this reason, most of the training set
do not have significantly measured parallaxes and HOGG percent even have
negative parallaxes!
But we will perform our training such that it will be unbiased in these
circumstances.
Indeed it would be necessarily biased if ever we did cut on parallax or
parallax signal-to-noise.

In detail,
for each star $n$ in the Training Sets, we take from \gaia\ \acronym{DR2} (\citealt{gaiadr2})
the astrometric parallax $\gparallax_n$ and its uncertainty $\sigma_n$.
Importantly, to each \gaia\ parallax measurement we add a positive
offset $\varpi_0$ to adjust for the under-estimation of
parallaxes reported by the \gaia\ Collaboration (\citealt{lindegren}).
The \gaia\ recommended value of $\varpi_0=0.029$\,mas (\citealt{lindegren}),
but we adopt a value of $\varpi_0 = 0.048$\,mas because that value
optimizes our cross-validation likelihood.
The spectrophotometric parallaxes do not depend strongly on this choice, except
for small differences at very small parallax (large distance).
The Training Sets are shown in \figurename~\ref{fig:training}.
\begin{figure}
\includegraphics[width=\textwidth]{./training_sample.pdf}
\caption{Color--color diagrams for the Training Set stars.
  Note that there is less color range in the Training Set than in the
  Parent Sample.\label{fig:training}}
\end{figure}

@@

For every star $n$ in the full Parent Sample we construct the feature
vector $x_n$ as
\begin{eqnarray}
x_n\T &\equiv& [1, G_n, \BP_n, \RP_n, J_n, H_n, K_n, W_{1n}, W_{2n}, \ln f_{1n}, \ln f_{2n}, \cdots, \ln f_{Ln}]
\quad ,
\end{eqnarray}
where the 1 permits a linear offset,
the photometry is from \gaia, \zmass, and \wise, respectively,
the fluxes in the $L=7405$ \apogee\ spectral pixels (for which there are reliably
and consistently data) for star $n$ are denoted
$f_{1n}$, $f_{2n}$, and so on,
and we have taken logarithms of those fluxes.
These feature vectors live in a $D$-dimensional space where $D=7414$
For every star $n$ in either of the Training Sets we additionally require---in
addition to these feature vectors---a \gaia-measured astrometric parallax $\gparallax_n$
and uncertainty $\sigma_n$.

@@

We optimize two models, one for Training Set A and one for Training Set B.
The two models are applied to the two splits of the Parent Sample, Parent Sample
A and Parent Sample B, using the A-trained model on Parent Sample B and
\foreign{vice versa}.
For the purposes of assessing the accuracy and precision of the model, this
train-and-test framework constitutes a two-fold cross-validation.
Results of this cross-validation is shown in \figurename~\ref{fig:xval}.
For the stars with highest \gaia-measured parallax signal-to-noise,
the spectrophotometric model is predicting the \gaia\ parallaxes with little bias
and a scatter of better than 9~percent.
\begin{figure}
\includegraphics[width=\textwidth]{residuals.pdf}
\caption{The predictive accuracy of the model, in the two-fold cross-validation.
  Each panel shows \gaia\ astrometric parallaxes $\gparallax_n$ on the vertical axis
  and our spectrophotometric parallaxes $\sparallax$ on the horizontal axis.
  The left panel shows the full Parent Sample, the middle panel shows the Training Set
  stars (which pass \gaia\ quality checks; see text), and the right panel shows a
  subset of stars with very high astrometric parallax signal-to-noise. This latter
  set plays no particular role in the method, but it can be used to demonstrate or
  assess the prediction precision. Fractional precision of the prediction
  is XXXX HOGG MEASURED HOW?\label{fig:xval}}
\end{figure}

We used this cross-validation framework to adjust the regularization parameters.
We (arbitrarily) chose zero regularization on the photometric components of the
$x_n$ vectors, and uniform, identical regularization strength for all of the
spectral (\apogee-pixel) components.
We did a coarse grid search in the value for the spectral-component value of the
$\lambda$ vector, using the A/B split as the two-fold cross-validation.
The 9-percent scatter is what we obtain at the optimal setting of the regularization
strength.

>>>>

% To-do items
% -----------
% - Coryn's point that we should carefully explain how our model is geometric.
% - check with WISE experts that we have the correct WISE photometry
% - write words about other linear models we have made or could make, probably in the discusssion.
% - are we citing all the proper DR2 papers we need to cite?

@@

\title{\textbf{%
Spectrophotometric parallaxes with linear models:\\
Accurate distances for luminous red-giant stars
}}

@@

\begin{abstract}\noindent
With contemporary infrared spectroscopic surveys like \apogee,
red-giant stars can be observed to distances and extinctions
at which geometric parallaxes are not precisely measured by \gaia.
Here we use the \apogee--\gaia\ overlap to train a purely linear model of
continuum-normalized \apogee\ spectroscopy
(plus \gaia, \zmass, and \wise\ photometry)
that predicts distance modulus (logarithmic parallax or distance)
for red stars with $0<\logg<2.2$.
The training makes use of a justified likelihood function for the \gaia\ parallaxes,
so that we never take the inverse or logarithm of a parallax, and so that
the training can be performed with full use of uncertainties and
without any cuts on parallax or parallax signal-to-noise ratio.
There are no physical inputs to the model; it is a pure regression designed
to accurately predict astrometric parallaxes.
The model includes an L1 regularization that zeros out the contributions of
most spectral pixels to the parallax estimates.
The training is performed
with leave-out subsamples such that no star's astrometry is used even indirectly in its
spectrophotometric parallax estimate.
The model is flexible enough---and has enough photometric information---to
correct for extinction by dust without any
input of explicit extinction or reddening estimates.
We re-label each star in the sample with a new spectrophotometric parallax;
these new parallaxes have uncertainties of a few to 15 percent, depending on data quality,
which is more precise than the \gaia\ parallax for the vast majority of targets.
These distance estimates open up great opportunities for
mapping the Milky Way disk with the next
generation of spectroscopic surveys, and especially \sdssv.
\end{abstract}

@@

If we want to make precise kinematic and element-abundance maps
of the Milky Way disk out to large heliocentric radii,
and especially through the extinction to the other side of the Galactic Center, we will
need to make use of luminous red giants, and we will need to observe in
the infrared.
These arguments motivate the \apogee\ (\citealt{apogee}),
\apogee-2, and \sdssv\ (\citealt{sdssv})
projects, which take spectra in the infrared, and deliver detailed abundances
along the entire red-giant branch up to the tip.
The maps made with these data will reveal the dynamics of the Milky Way and its disk,
and show us how and where stars form, and how they migrate around (and out of) the
disk with time.

As we operate these spectroscopic surveys,
the \gaia\ Mission (\citealt{gaia}) has been revolutionizing our view of the disk.
It delivers good proper motions everywhere,
and extremely valuable parallax information
locally.
However, it does not deliver precise geometric parallaxes over a large fraction of
the disk, and certainly not in the dusty and crowded regions.
For these reasons, the value of \gaia\ in these infrared spectroscopy projects is not
to deliver distance information directly, but rather to calibrate stellar models,
which then deliver distance information through relationships between stellar
luminosities and spectral characteristics.

Luminous red-giant stars are valuable for mapping the Milky Way disk for a number
lf reasons, only one of which is their great luminosities (and hence brightnesses
even at large distance).
They are extremely common (frequent) stars, much more common than distance-indicating
stars like Cepheids.
They are produced by stellar populations at all metallicities and almost all ages,
so they can be found in all parts of the Galaxy.
They have temperatures and surface gravities such that it is
straightforward to spectroscopically measure metallicities,
detailed chemical abundances, and stellar parameters, even including ages (\citealt{martig, nessage}).
The red-giant branch is photometrically near-orthogonal to reddening vectors by dust,
so they are easy to de-redden or dust-correct.
And finally, in physical models of red giants, their predicted luminosities are simple
functions of composition, surface gravity, temperature, and age.
Thus, if we can measure these properties of red giants, we can (in principle) predict
their luminosities and use them as standard candles.
That is---like type Ia supernovae---red giants are standardizable candles.

In this project, we develop purely data-driven techniques to predict
red-giant parallaxes or distances with spectroscopy and photometry.
Our method capitalizes on the point that red-giant stars are dust-correctable,
standardizable candles, but it makes no use whatsoever of any stellar interior,
stellar exterior, or dust-reddening models.
We only use these physical expectations about red giants to recommend
data features---inputs---for a completely data-driven model.
The data-driven method uses patterns in the observed data themselves
to discover the relationships between spectral features in the spectra of the stars,
photometry (including colors), and parallax (or distance).

@@

Simple data-driven models (like the ones we will employ here)
also suffer from many limitations for this kind of work:
Because they are constrained only by the data, and not physical law,
they can learn relationships that we know
to be physically unlikely, and (by the same token) they can't benefit from our
physical knowledge.
They are required, in some sense, to spend some of the information in the data on
learning physical laws or relationships.
Not all of the information in the data is being used
optimally when we ignore our physical models.
Relatedly, they are not (usually) interpretable or generalizable:
A data-driven model trained with one kind of data can not usually be used in another
context with very different input data.
And the internal properties of the model are not useful, in general, for improving
our understanding or intuition about the astrophysics in play.

@@

What follows is a regression:
We use \gaia\ measurements of parallax to learn relationships between 
the spectroscopic and photometric properties of stars and their parallaxes.
We then use that learned model to estimate improved parallaxes for stars with
poor (or no) \gaia\ measurements.
This regression is very sensitive to certain aspects of the data or experimental
design.
One is that we will do far better with the regression if we have an accurate
model for noise process in the \gaia\ data.
For this reason, we make use of a justifiable likelihood function for the
\gaia\ astrometry (\citealt{gaialf}).
Another is that regressions can be strongly biased if we have a bad selection function
or data censoring.
This means that we cannot apply cuts to the \gaia\ astrometry or signal-to-noise
unless those cuts are correctly accounted for in our generative model for the
data sample.
Below, we will make no such cuts. That means that our training step has the odd
property that much of the training data is low in signal-to-noise and some of the
training parallax measurements are even negative!
The fact that we are using a justified likelihood function ensures that these
data will only affect the model in sensible and justifiable ways.
The alternative---cutting to high signal-to-noise data---is the more traditional
approach, but it leads to biased inferences (unless the cuts are included correctly
in the likelihood function).
The method in which we make no cuts and keep an umodified, justified likelihood function
is unbiased, and simple.

@@

Our position is that a methodological technique is correct inasmuch as
it delivers correct or best results \emph{under its particular assumptions}.
In that spirit, we present here the assumptions of the method
explicitly.
\begin{description}
\item[features] Perhaps our most fundamental assumption is that the parallax
of a star can be predicted from the features we provide, which are
the full set of (pseudo-continuum-normalized) pixels from the \apogee\ (\citealt{apogee}) spectrum,
plus the $G$, $\BP$, and $\RP$ photometry from \gaia\ (\citealt{gaia}),
plus the $J$, $H$, and $K_s$ photometry from \zmass\ (\citealt{zmass}),
plus the $W_1$ and $W_2$ photometry from \wise\ (\citealt{wise}).
That is, we assume that these spectrophotometric \emph{features} are sufficient
to predict the parallax in the face of variations in stellar
age, evolutionary phase, composition, and other parameters, and also interstellar
extinction.

@@

\item[linearity] Perhaps the most restrictive assumption of this work is that
the logarithmic parallax (or, equivalently, the distance modulus) 
can be predicted as a \emph{completely linear function} of
the chosen features. We are only assuming this linearity in a small range of stellar
surface gravity, but this assumption is strong, and limits strongly the
flexibility or capacity of the model.
We make it to ensure that our method is easy to optimize, and the results are easy
to interpret.

\item[likelihood] We assume that the \gaia\ parallaxes are generated by a particular
stochastic process, in which the difference between the \gaia-Catalog parallax (plus
a small \gaia-recommended offset; \citealt{lindegren}) and the true parallax is effectively drawn from a
Gaussian with a width set by the \gaia-Catalog uncertainty on the parallax.
This is the standard assumption in all properly probabilistic \gaia-Mission inferences
to date, but it subsumes a number of related assumptions, like that the \gaia\ noise
model is correct, that the stars are only covariant with one another at negligible levels, and that
there are no significant outliers or large-scale systematics.
\end{description}

@@

In addition, we assume that many entries in the parameter $D$-vector $\theta$ will be zero
(the ``sparsity'' assumption).
In order to encourage this, we optimize not $\chi^2$ but a regularized objective function
\begin{eqnarray}
\hat{\theta} &\leftarrow& \argmin_{\theta}\left[\frac{1}{2}\,\chi^2(\theta) + \lambda\,||P\cdot\theta||_1^1\right]
\label{eq:objective}\quad ,
\end{eqnarray}
where
$\lambda$ is a regularization parameter,
$P$ is a projection operator that selects out from the $\theta$ vector only those components
that pertain to the \apogee\ spectral pixels,
and $||y||_1^1$ is the L1-norm or sum of absolute values of the components of $y$.
This kind of regularization (L1) adds a convex term to the optimization and leads to
sparse optima.
The $P$ operator makes it such that we only regularize the components of $\theta$ that multiply
the spectral pixels; we ask for the spectral model to be sparse but we don't ask for the photometric
model to be sparse.
We set the value of $\lambda$ by cross-validation.

This optimization is not convex---there are multiple optima in general.
In particular, there is a large, degenerate, pathological optimum where
the exponential function underflows, all parallaxes are predicted to vanish,
and there is no gradient of anything with respect to the parameter $D$-vector $\theta$.
This large, bad optimum (it is a local, not global, optimum) must be avoided in optimization.
In practice we avoid it by optimizing first for the very highest signal-to-noise
(in astrometric parallax) Training Set stars, and then use that first optimum as a
good first guess or initialization for the full optimization.
This method works because in the limit of high signal-to-noise, the problem asymptotically approaches
the convex problem of L1-regularized linear least-square fitting.
Optimization is performed with \code{scipy.optimize} using the
\acronym{L-BFGS-B} algorithm (HOGG CITE).

Once the model is optimized, the output of the model is a prediction of the parallax,
or really what we will call the \emph{spectrophotometric parallax}.
This spectrophotometric parallax
$\sparallax_m$ for any star $m$ in the validation or test set is
assigned according to
\begin{eqnarray}
\sparallax_m &\leftarrow& \exp(\hat{\theta}\cdot x_m)
\quad ,
\end{eqnarray}
where
$\hat{\theta}$ is the optimal parameter vector according
to \equationname~(\ref{eq:objective}),
and
$x_m$ is the feature $D$-vector for star $m$.
Thus the model can be trained on a training set of stars and applied to
a validation set or a test set to make predictions.
The only requirements are that every test-set object $m$  must have a full feature
vector $x_m$ just as every training-set object $n$ must have a full feature
vector $x_n$.

One final detail: We perform all of the fitting in a two-fold train-and-test framework,
in which the data are split into two disjoint subsets, A and B.
The model trained on the A data is used to predict or produce
the spectrophotometric parallaxes $\sparallax_m$ values (and hence
the distances and parallaxes) of the B data,
and the model trained on the B data is used to 
produce the $\sparallax_m$ values of the A data.
This ensures that, in the estimate of any individal star's
spectrophotometric parallax, none of the \gaia\ data pertaining to that star were used.
This makes the parallax estimates from the spectrophotometric feature vectors
$x_m$ statistically independent of the \gaia\ parallax measurements.
They are not independent globally---\gaia\ data were used to train the model---but
each star's spectrophotometric parallax estimate is independent
of the astrometric estimate from \gaia\ on a star-by-star basis.

@@

We need complete feature-vector information for all stars.  For this
reason, we define the Parent Sample to be the set of all stars that
meet the \apogee\ and \gaia\ cuts and also
have the complete set of photometry: $G_n$, $\BP_n$, $\RP_n$, $J_n$,
$H_n$, $K_n$, $W_{1n}$, and $W_{2n}$ and also an \apogee\ spectrum.
In addition, we applied two light color cuts to remove stars with obviously
contaminated or outlying photometry:
\begin{eqnarray}
(J_n - K_n) &<& (+0.4\,\mathrm{mag}) + 0.45 * (\BP_n - \RP_n)
\\
(H_n - W_{2n}) &>& (-0.05\,\mathrm{mag})
\quad .
\end{eqnarray}
These cuts removed roughly 2~percent of the \apogee\ stars.
This Parent Sample contains 44784 stars and is shown in \figurename~\ref{fig:parent}.
\begin{figure}
\includegraphics[width=\textwidth]{parent_sample.pdf}
\caption{Color--color diagrams for the 44784 Parent Sample stars.
  The points are colored by
  the \apogee\ pipeline estimates of surface gravity.
  The color cuts used to trim outliers are shown as grey lines.\label{fig:parent}}
\end{figure}

Every Parent-Sample star gets, in addition, a randomly assigned binary
label (A or B).
This is used for two-fold validation and jackknife.
In short, we will
use the model trained on the Training Set A data to assign spectrophotometric parallaxes
to the Parent Sample B data and
use the model trained on the Training Set B data to assign spectrophotometric parallaxes
to the Parent Sample A data.

For training, we need A and B Training Sets.
We define the Training Set stars to be Parent-Sample stars that also
have a measured \gaia\ parallax.
The Training-Set stars, in addition to meeting all Parent-Sample cuts,
also must meet ACE WHAT ADDITIONAL CUTS to ensure that the parallax
measurements are good.
These cuts leave 28226 stars total in the Training Sets.
But---as we have emphasized above---we do not cut ever on parallax or
parallax error or their ratio. For this reason, most of the training set
do not have significantly measured parallaxes and more than 2~percent even have
negative parallaxes!
But we will perform our training such that it will be unbiased in these
circumstances.
Indeed it would be necessarily biased if ever we did cut on parallax or
parallax signal-to-noise.

In detail,
for each star $n$ in the Training Sets, we take from \gaia\ \acronym{DR2} (\citealt{gaiadr2})
the astrometric parallax $\gparallax_n$ and its uncertainty $\sigma_n$.
Importantly, to each \gaia\ parallax measurement we add a positive
offset $\varpi_0$ to adjust for the under-estimation of
parallaxes reported by the \gaia\ Collaboration (\citealt{lindegren}).
The \gaia\ recommended value is $\varpi_0=0.029$\,mas (\citealt{lindegren}),
but we adopt $\varpi_0 = 0.048$\,mas because that value
optimizes our cross-validation objective.
The spectrophotometric parallaxes do not depend strongly on this choice, except
for small differences at very small parallax (large distance).
The Training Sets are shown in \figurename~\ref{fig:training}.
\begin{figure}
\includegraphics[width=\textwidth]{./training_sample.pdf}
\caption{Color--color diagrams for the 28226 Training Set stars.
  Note that there is less color range in the Training Set than in the
  Parent Sample.\label{fig:training}}
\end{figure}

@@

For every star $n$ in the full Parent Sample we construct the feature
vector $x_n$ as
\begin{eqnarray}
x_n\T &\equiv& [1, G_n, \BP_n, \RP_n, J_n, H_n, K_n, W_{1n}, W_{2n}, \ln f_{1n}, \ln f_{2n}, \cdots, \ln f_{Ln}]
\quad ,
\end{eqnarray}
where the 1 permits a linear offset,
the photometry is from \gaia, \zmass, and \wise, respectively,
the fluxes in the $L=7405$ \apogee\ spectral pixels (for which there are reliably
and consistently data) for star $n$ are denoted
$f_{1n}$, $f_{2n}$, and so on,
and we have taken logarithms of those fluxes.
These feature vectors live in a $D$-dimensional space where $D=7414$.
For every star $n$ in either of the Training Sets we additionally require---in
addition to these feature vectors---a \gaia-measured astrometric parallax $\gparallax_n$
and uncertainty $\sigma_n$.

@@

We optimize two models, one for Training Set A and one for Training Set B.
The two models are applied to the two splits of the Parent Sample, Parent Sample
A and Parent Sample B, using the A-trained model on Parent Sample B and
\foreign{vice versa}.
For the purposes of assessing the accuracy and precision of the model, this
train-and-test framework constitutes a two-fold cross-validation.
Results of this cross-validation is shown in \figurename~\ref{fig:xval}.
For the stars with highest \gaia-measured parallax signal-to-noise,
the spectrophotometric model is predicting the \gaia\ parallaxes with little bias
and a scatter of less than 9~percent.
\begin{figure}
\includegraphics[width=\textwidth]{residuals.pdf}
\caption{The predictive accuracy of the model, in the two-fold cross-validation.
  Each panel shows \gaia\ astrometric parallaxes $\gparallax_n$ on the vertical axis
  and our spectrophotometric parallaxes $\sparallax$ on the horizontal axis.
  The left panel shows the full Parent Sample, the middle panel shows the Training Set
  stars (which pass \gaia\ quality checks; see text), and the right panel shows a
  subset of stars with very high astrometric parallax signal-to-noise. This latter
  set plays no particular role in the method, but it can be used to demonstrate or
  assess the prediction precision. The fractional precision of the prediction in the
  right panel is better than 9 percent.\label{fig:xval}}
\end{figure}

We used this cross-validation framework to adjust the regularization parameter $\Lambda$.
A coarse grid search in the value for the spectral-component value of the
$\lambda$ vector, using the A/B split as the two-fold cross-validation, with the objective
of the cross validation being prediction accuracy.
We get a prediction accuracy of better than 9 percent 
at the optimal setting of the regularization strength.
